### 2023-05-08 chapter0/chapter0.tex
Changed genetic -> biological because more generic
Implemented changes from Susan's comments (largely typos and clarifications)

Whereas hemoglobin is an example of a protein mediating transport, proteins are also involved in transmitting signals and catalyzing chemical reactions. For example, the back of the eye contains a light-sensitive surface called the retina which is composed of photoreceptor cells. These cells respond to light because they produce special proteins called opsins that translate light into chemical and electrical signals which are then interpreted by the brain. Humans, and primates broadly, have three types of opsins, which are sensitive to red, green, and blue light, respectively, that mediate our color vision. Color blindness is the result of photoreceptor cells missing one of these proteins, typically either the red or green opsin. In contrast to sickle cell disease, this condition is caused by a missing rather than a mutated protein. In other cases, however, a protein is not missing or mutated, but instead not produced at the right time and place. For example, lactose is a sugar found in milk that requires a specific protein, lactase, to metabolize properly. Many humans who can digest milk products in childhood lose this ability in adulthood because they stop producing lactose. As a result, lactose in dairy products passes undigested into the colon where it is broken down by bacteria, causing symptoms such as bloating and diarrhea.

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the 20 ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in turn a direct result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

@@

In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as conformational ensembles, \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably. Furthermore, IDRs populate a continuum of structural states over time, whereas when folded domains undergo large conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient. Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended coils to more compact globules. The specific conformational class of a given IDR, however, is dictated by its local composition of amino acid residues. Because the chemical properties of each amino acid in the protein alphabet are dictated its specific arrangement of atoms, each has a unique impact on a protein's structure. However, to simplify discussion and analysis amino acids are often compared by quantitative factors like size or charge. One of the most useful scales for describing an amino acid's overall effect on protein structure is hydrophobicity, which measures a molecule's tendency to associate with water. Molecules which attract water are hydrophilic (water loving), and molecules which repel water are hydrophobic (water fearing). Hydrophilic molecules, like sugar or alcohol, easily dissolve in water, whereas hydrophobic molecules, like fats and oils, remain separate. Since the cellular environment is largely water, the hydrophobic residues in proteins tend to aggregate into a hydrophobic core. This hydrophobic collapse is a major driving force in the early stages of protein folding, so a protein's relative number of hydrophilic and hydrophobic residues is a key determinant of whether it is folded or disordered.

@@

The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners. Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions. As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling. Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life. However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles. Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time. However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by interactions between proteins, which in turn create a change in the state of the system that propagates the signal. Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation. Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein. Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change. Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states. In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior. Because IDRs are generally exposed to their environment, they are frequent targets of PTMs. However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs). SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners. In contrast to the highly structured interfaces that characterize interactions with folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}. Thus, their interactions are relatively weak and highly transient. As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously. IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Buljan2012, Wright2014}. Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues. Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''

Some of the most prominent examples of IDRs in cell signaling are found in the proteins that regulate the creation of other proteins. A protein's amino acid sequence is encoded in DNA in a unit called a gene, and since all cells in an organism share the same genome, they can in principle synthesize any protein encoded in it.\footnote{The DNA sequences of different cells in an organism are not always strictly identical. For example, spontaneous mutations can create changes that range from single-letter substitutions to large-scale rearrangements. In other cases, such as during the creation of sex or immune cells, a cell's DNA is intentionally modified to generate genetic diversity.} However, different cell types instead express unique complements of proteins that in large part determine their identities. This selective conversion of the information stored in DNA into proteins is called gene expression and is a highly regulated process that varies in space and time, \textit{i.e.} between different cells and within a single cell. Though gene expression involves dozens, if not hundreds, of distinct biochemical reactions, most occur as steps within two major processes: transcription and translation. In transcription, the sequence of a protein encoded the cell's DNA is transcribed into an intermediate molecule called messenger RNA (mRNA). mRNA is also a kind of nucleic acid, and is chemically closely related to DNA. However, as a much smaller molecule containing only the information needed to synthesize a protein, mRNA is more easily transported and manipulated in subsequent steps. Thus, if a gene stored as DNA is a cell's master record of a protein sequence, mRNA is its working copy. In translation, a vast molecular machine called the ribosome then synthesizes a protein by translating the information encoded in a strand of mRNA into a sequence of amino acids.

@@

\section{Aims}
Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between the sequence and function remains poorly understood. In contrast, predictions of structure and function are readily available for many folded domains. Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly. Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}. Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone was an active area of research for decades. Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences. Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences. Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.

@@

These analyses were conducted using IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research. However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems. As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation. Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in gene regulation as well as human health and disease. The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution. For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast. In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.

@@

These comparisons, however, require the identification of IDRs with common ancestry that perform equivalent functions across many distinct organisms. Given the difficulties with identifying similar IDRs by their sequences, this may seem like a chicken and egg problem. Fortunately, IDRs are frequently associated with more conserved folded domains. Thus, identifying evolutionarily related proteins by their overall sequence signatures and aligning them will in turn identify the equivalent IDRs in those sequences. The first step of an evolutionary analysis of IDRs, then, is the identification of proteins with common ancestry, called orthologs. Since the first genomes were sequenced in the late 1990s, researchers have developed techniques for identifying and aligning orthologs~\cite{Fleischmann1995, Goffeau1996, CESC1998, Tatusov1997}. While these methods are generally effective, they are conducted by automated computational pipelines and prone to errors when processing the highly divergent sequences that characterize many IDRs. The evolutionary relationships between the genomes of closely related species generally make such mistakes easier to identify, and fortunately over the past five years advances in DNA sequencing technology have yielded dramatic increases in the number of sequenced genomes in the \textit{Drosophila} genus. However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leverage such genomic redundancy to minimize errors. Thus, in the first chapter I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B). In the second chapter I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to dissect the forces driving their evolution and detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F). Finally, in the third chapter I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.

>>>>

Whereas hemoglobin is an example of a protein mediating transport, proteins are also involved in transmitting signals and catalyzing chemical reactions. For example, the back of the eye contains a light-sensitive surface called the retina which is composed of photoreceptor cells. These cells respond to light because they produce special proteins called opsins that translate light into chemical and electrical signals which are then interpreted by the brain. Humans, and primates broadly, have three types of opsins, which are sensitive to red, green, and blue light, respectively, that mediate our color vision. Color blindness is the result of photoreceptor cells missing one of these proteins, typically either the red or green opsin. In contrast to sickle cell disease, this condition is caused by a missing rather than a mutated protein. In other cases, however, a protein is not missing or mutated, but instead not produced at the right time and place. For example, lactose is a sugar found in milk that requires a specific protein, lactase, to metabolize properly. Many humans who can digest milk products in childhood lose this ability in adulthood because they stop producing lactase. As a result, lactose in dairy products passes undigested into the colon where it is broken down by bacteria, causing symptoms such as bloating and diarrhea.

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the 20 ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary ``receptor'' and ``donor'' sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in part a result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

@@

In contrast, IDRs have no fixed spatial relationship between their atoms, so their structures are sometimes described as conformational ensembles, \textit{i.e.} collections of conformations where the distances and orientations between residues can vary considerably. Furthermore, IDRs populate a continuum of structural states over time, whereas when folded domains undergo large conformational changes, they are usually triggered by specific environmental signals or chemical modifications, and the intermediate structures are transient. Despite the diversity of conformations available to IDRs, they can be broadly grouped into one of several qualitative descriptions which range from extended coils to more compact globules. The specific conformational class of a given IDR, however, is dictated by its local composition of amino acid residues. Because the chemical properties of each amino acid in the protein alphabet are dictated its specific arrangement of atoms, each has a unique impact on a protein's structure. However, to simplify discussion and analysis amino acids are often compared by quantitative factors like size or charge. One of the most useful scales for describing an amino acid's overall effect on protein structure is hydrophobicity, which measures a molecule's tendency to associate with water. Molecules which interact with water are hydrophilic (water loving), and molecules which do not are hydrophobic (water fearing). Thus, hydrophilic molecules, like sugar or alcohol, easily dissolve in water, whereas hydrophobic molecules, like fats and oils, remain separate. Since the cellular environment is largely water, the hydrophobic residues in proteins tend to cluster into a hydrophobic core. This hydrophobic collapse is a major driving force in the early stages of protein folding, so a protein's relative number of hydrophilic and hydrophobic residues is a key determinant of whether it is folded or disordered.

@@

The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners. Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions. As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling. Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life. However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles. Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time. However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by interactions between proteins, which in turn create a change in the state of the system that propagates the signal. Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation. Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein. Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change. Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states. In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior. Because IDRs are generally exposed to their environment, they are frequent targets of PTMs. However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs). SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners. In contrast to the highly structured interfaces that characterize interactions with folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}. Thus, their interactions are relatively weak and highly transient. As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously. IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Buljan2012, Wright2014}. Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues. Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different biological ``programs.''

Some of the most prominent examples of IDRs in cell signaling are found in the proteins that regulate the creation of other proteins. A protein's amino acid sequence is encoded in DNA in a unit called a gene, and since all cells in an organism share the same genome, they can in principle synthesize any protein encoded in it.\footnote{The DNA sequences of different cells in an organism are not always strictly identical. For example, spontaneous mutations can create changes that range from single-letter substitutions to large-scale rearrangements. In other cases, such as during the creation of sex or immune cells, a cell's DNA is intentionally modified to generate genetic diversity.} However, different cell types instead express unique complements of proteins that in large part determine their identities. This selective conversion of the information stored in DNA into proteins is called gene expression and is a highly regulated process that varies in space and time, \textit{i.e.} between different cells and within a single cell. Though gene expression involves dozens, if not hundreds, of distinct biochemical reactions, most occur as steps within two major processes: transcription and translation. In transcription, the sequence of a protein encoded in the cell's DNA is transcribed into an intermediate molecule called messenger RNA (mRNA). mRNA is also a kind of nucleic acid, and is chemically closely related to DNA. However, as a much smaller molecule containing only the information needed to synthesize a protein, mRNA is more easily transported and manipulated in subsequent steps. Thus, if a gene stored as DNA is a cell's master record of a protein sequence, mRNA is its working copy. In translation, a vast molecular machine called the ribosome then synthesizes a protein by translating the information encoded in a strand of mRNA into a sequence of amino acids.

@@

\section{Aims}
Despite recent advances in identifying IDRs and their conformational ensembles from their sequences alone~\cite{Mao2010, vanderLee2014, Das2015, Necci2021}, the relationship between their sequence and function remains poorly understood. In contrast, predictions of structure and function are readily available for many folded domains. Because they make specific contacts between residues, the sequences of folded domains are generally conserved or evolve slowly. Thus, the specific sequence of a folded domain constitutes its unique ``signature.'' Traditional bioinformatics techniques use these signatures to detect similar protein sequences and transfer structural and functional information between them~\cite{Camacho2009, Eddy2009, Mistry2020}. Though this approach is simple in principle, it is extremely difficult to perfect in practice, and fully mapping the relationship between the sequence and structure of folded domains alone hsa been an active area of research for decades. Part of the challenge is the available data is extremely sparse relative to the sheer number of possible protein sequences. Even for a moderately sized protein of 100 amino acid residues, there are $20^{100} \approx 1.3 \times 10^{130}$ possible sequences. Only in the past few years have researchers in many senses solved this problem by using advanced techniques from machine learning to leverage the information encoded in nearly two hundred thousand experimentally determined structures~\cite{Jumper2021}.

@@

These analyses were conducted using IDRs identified in various species of yeast, which is a widely used model organism in molecular biology research. However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems. As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation. Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in gene regulation as well as human health and disease. The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution. For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are a property of IDRs which is unique to yeast. In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.

@@

These comparisons, however, require the identification of IDRs with common ancestry that perform equivalent functions across many distinct organisms. Given the difficulties with identifying similar IDRs by their sequences, this may seem like a chicken and egg problem. Fortunately, IDRs are frequently associated with more conserved folded domains. Thus, identifying evolutionarily related proteins by their overall sequence signatures and aligning them will in turn identify the equivalent IDRs in those sequences. The first step of an evolutionary analysis of IDRs, then, is the identification of proteins with common ancestry, called orthologs. Since the first genomes were sequenced in the late 1990s, researchers have developed techniques for identifying and aligning orthologs~\cite{Fleischmann1995, Goffeau1996, CESC1998, Tatusov1997}. While these methods are generally effective, they are conducted by automated computational pipelines and prone to errors when processing the highly divergent sequences that characterize many IDRs. The evolutionary relationships between the genomes of closely related species generally make such mistakes easier to identify, and fortunately over the past five years advances in DNA sequencing technology have yielded dramatic increases in the number of sequenced genomes in the \textit{Drosophila} genus. However, because the existing methods for ortholog identification were designed for fewer or more distantly related genomes, they do not fully leverage such genomic redundancy to minimize errors. Thus, in the first chapter, I develop a novel method for identifying orthologs which addresses this shortcoming and apply it to 33 \textit{Drosophila} genomes to generate a set of aligned orthologs (Fig.~\ref{fig:aims_overview}A-B). In the second chapter, I then identify rapidly evolving IDRs in these alignments and analyse them with a variety of evolutionary models to dissect the forces driving their evolution and detect patterns of conservation (Fig.~\ref{fig:aims_overview}C-F). Finally, in the third chapter, I discuss several software tools and tutorials for fitting statistical models to data, which were created while pursuing the previous aims.
